When Do Repeat Yourself?\\ Voices from the Trenches of Linux Kernel Maintainers on Code Duplication

### 2025-10-24 paper/techdebt2026/techdebt.tex
remove mitigation 1 and rephase takeway 1 a little
techdebt2026: techdebt: revise introduction
Make adjustments to the introduction. Among which:
- Add the DRY quote
- Make the introduction more concise
- Give more weight to the contrasting interactions with the maintainers
  in terms of deduplications.

Rename the highlight box macro also, as a collateral effect.

Signed-off-by: David Tadokoro <davidbtadokoro@usp.br>

\newenvironment{takeaway}[1]{%
  \begin{tcolorbox}
  \textbf{#1:} \itshape}{\end{tcolorbox}}
%% Takeaway Boxes %%

@@

Coordinating hundreds or thousands of developers working on a software product is complex. Creating similar or even completely duplicated code is a frequent and often unavoidable outcome. Excessive duplication can negatively impact the maintainability of a project. When duplications occur, any change to a repeated code segment typically requires manual replication across all other copies. Developers must, therefore, remain constantly vigilant to synchronize different parts of the codebase, which is a highly error-prone and seemingly simple responsibility. With many duplications, modifications become slow, complex, and fragile. Removing or refactoring duplicated code demands great care to preserve the original behavior. Even minor mistakes can introduce subtle state changes that may lead to serious problems in the future.

The Linux kernel is a foundational Free/Libre and Open Source Software (FLOSS) project, a vital component of the world's digital infrastructure. Its maintenance is a massive undertaking, encompassing over 28 million lines of code and involving contributions from over twenty thousand developers. Within this context, code duplication remains a significant challenge,  reducing the code readability and increasing the risk of introducing bugs~\cite{harmone,harmtwo}. This issue is particularly prominent in kernel device drivers, which account for over 66\% of the Linux kernel source code.

%TODO: temos que citar o conceito de DRY aqui ... (Book: Pragmatic Programmer, The: From Journeyman to Master)
% We feel that the only way to develop software reliably, and to make our developments easier to understand and maintain, is to follow what we call the DRY principle: Every piece of knowledge must have a single, unambiguous, authoritative representation within a system.
The detection of code duplication, or \textit{code clones}, has been a research topic for decades~\cite{firstman}. The literature offers a widely accepted taxonomy that classifies clones into four types based on their similarity, from exact copies (Type-1) to semantically equivalent but syntactically different fragments (Type-4)~\cite{litreview}. Several detection techniques have emerged, including textual, token-based, tree-based, and graph-based approaches~\cite{litreview}. These efforts have culminated in state-of-the-art techniques, such as the graph-based approach proposed by Liu et al.~\cite{tailor}.

To address the code duplication issue, we developed a command-line tool designed to support Linux kernel maintenance by detecting and analyzing function-level duplications. This paper presents a multimethod ethnographic study to validate the tool capabilities in identifying duplications and supporting contributions to the Linux kernel by mitigating redundant code. Therefore, the study investigates how the Linux kernel community perceives the code quality related to code clones. Our findings indicate that maintainers consider not only duplication but also factors such as readability and appropriate levels of abstraction regarding code maintainability.

Ethnography is a well-established qualitative research method used to understand people, their cultures, and their work practices~\cite{bookethno}. It enables researchers to gain insights into values, beliefs, and practices from the perspective of community members~\cite{ethnosoft}.  
In this study, we first adopted a complete participant observation approach, where the main author actively engaged in mitigating code duplications while documenting the process and outcomes. Later, we conducted a partipant-as-observer study involving students tasked with resolving duplications identified by the tool. These combined studies showed that our tool can collaborate to reduce the entry barrier for new contributors. The student groups (and we) used the tool to guide their initial contributions to the Linux kernel.

@@

\begin{takeaway}{Takeaway T2 (Readability)}
  Maintainers notably prioritized readability over removing duplications in
  many different contexts. In their view, going against the \textbf{DRY
  principle} was worth it if it resulted in code that is easier to read and
  understand.
\end{takeaway}

@@

\begin{takeaway}{Takeaway T3 (Integration Overhead)}
  A maintainer suggested that the effort put into the contribution after it was
  accepted (i.e., the process of propagating the change to upper maintainers)
  was considered. In other words, maintainers may \textbf{evaluate the
  trade-off between the impact of contributions (not only deduplications) and
  the struggle to integrate them into the complex Linux project}.
\end{takeaway}

@@

%TODO Na citação abaixo, trazida pelo Arthur, pensar se não vale colocar as
%referências base do paper do Lanza: ``Developers need to mentally reconstruct
%the role and behavior of a class from scattered fragments of code [8],
%presented as one page of text at a time. This process is cognitively demanding
%[9], slow, and error-prone, especially in large systems with many
%interdependent components [10].''
%
% Referências do paper do Lanza para esse trecho:
%[8] M.-A. Storey, “Theories, methods and tools in program comprehension:
%Past, present and future,” in Proceedings of IWPC 2005. IEEE, 2005,
%pp. 181–191.
%[9] J. Sillito, G. C. Murphy, and K. De Volder, “Questions programmers
%ask during software evolution tasks,” in Proceedings FSE 2006. ACM,
%2006, pp. 23––34.
%[10] G. C. Murphy, D. Notkin, and K. Sullivan, “Software reflexion models:
%Bridging the gap between source and high-level models,” in Proceedings
%of FSE 1995. ACM, 1995, pp. 18–28.
\begin{takeaway}{Takeaway T4 (Cognitive Load)}
  Duplications that made it possible to read the code without needing to jump
  to the function definition (to see its behavior) or the function call (to see
  its concrete parameters) were justifiable to reduce cognitive load when
  parsing through the code~\cite{skylines-cognitive-load}.
\end{takeaway}

@@

\begin{takeaway}{Takeaway T5 (Performance)}
  Given the low-level development context of Linux and the IIO subsystem, which
  is usually related to devices with limited resources, we detected hints that
  duplication may be forgivable if it improves performance.
\end{takeaway}

>>>>

\newenvironment{highlight-box}[1]{%
  \begin{tcolorbox}
  \textbf{#1:} \itshape}{\end{tcolorbox}}
%% Takeaway Boxes %%

@@

Hundreds or thousands of developers working together on a software product is a complex and demanding task. Similar or even exact copies are a frequent and often unavoidable outcome. Excessive duplication can negatively impact the maintainability of a project, as, when it occurs, any change to a repeated code segment (to fix a bug, for instance) requires replication across all other copies. Developers must, therefore, remain constantly vigilant to synchronize different parts of the codebase, which is a highly error-prone yet seemingly simple responsibility. Regarded as a major source of technical debt, removing or refactoring duplicated code demands great care to preserve the original behavior, with minor mistakes possibly leading to serious problems in the future.

A widespread dogma among developers is the \textit{Don't Repeat Yourself} principle~\cite{pragmatic-programmer}, stated as follows.

\begin{highlight-box}{DRY principle}
  ``We feel that the only way to develop software reliably, and to make our
  developments easier to understand and maintain, is to follow what we call the
  DRY principle: Every piece of knowledge must have a single, unambiguous,
  authoritative representation within a system''.
\end{highlight-box}

The Linux kernel is a foundational Free/Libre and Open Source Software (FLOSS) project, a vital component of the world's digital infrastructure. Considering only lines of code (no comments or blank lines) of strictly programming languages, in version 6.17, the main repository has more than 27 million lines of code and involved contributions from over twenty thousand developers. Within this context, code duplication remains a significant challenge,  reducing the code readability and increasing the risk of introducing bugs~\cite{harmone,harmtwo}. This issue is particularly prominent in kernel device drivers, which account for over 66\% of the Linux kernel source code.

The detection of code duplication, or \textit{code clones}, has been a research topic for decades~\cite{firstman}. The literature offers a widely accepted taxonomy that classifies clones into four types based on their similarity, from exact copies (Type-1) to semantically equivalent but syntactically different fragments (Type-4)~\cite{litreview}. Several detection techniques have emerged, including textual, token-based, tree-based, and graph-based approaches~\cite{litreview}. These efforts have culminated in state-of-the-art techniques, such as the graph-based approach proposed by Liu et al.~\cite{tailor}.

To address code duplication in the Linux kernel, we developed a command-line tool capable of detecting and analyzing function-level duplications. We conducted a multimethod ethnographic study to test the tool capabilities by identifying duplications and submitting contributions, directly or indirectly, to the Linux kernel, removing the redundant code. Through interactions with maintainers, this study gained insights into how the Linux community perceives the code quality related to code clones. Contrary to the DRY principle, our findings indicate that, in some contexts, maintainers can tolerate duplications for better readability, performance increases, or even to avoid integration overhead.\looseness=-1

Ethnography is a well-established qualitative method for understanding people, their cultures, and work practices~\cite{bookethno}. It provides insights into community members’ values, beliefs, and practices~\cite{ethnosoft}. In this study, we first employed a complete participant observation approach, actively contributing by performing deduplications while documenting the process. We then conducted a participant-as-observer study with students resolving duplications identified by the tool. Beyond demonstrating that our tool lowered the entry barrier for new contributors, these studies yielded intriguing observations on kernel deduplications. This paper describes the aforementioned studies, along with our visions and reflections from diving into the development process.

@@

\begin{highlight-box}{Takeaway T2 (Readability)}
  Maintainers notably prioritized readability over removing duplications in
  many different contexts. In their view, going against the \textbf{DRY
  principle} was worth it if it resulted in code that is easier to read and
  understand.
\end{highlight-box}

@@

\begin{highlight-box}{Takeaway T3 (Integration Overhead)}
  A maintainer suggested that the effort put into the contribution after it was
  accepted (i.e., the process of propagating the change to upper maintainers)
  was considered. In other words, maintainers may \textbf{evaluate the
  trade-off between the impact of contributions (not only deduplications) and
  the struggle to integrate them into the complex Linux project}.
\end{highlight-box}

@@

%TODO Na citação abaixo, trazida pelo Arthur, pensar se não vale colocar as
%referências base do paper do Lanza: ``Developers need to mentally reconstruct
%the role and behavior of a class from scattered fragments of code [8],
%presented as one page of text at a time. This process is cognitively demanding
%[9], slow, and error-prone, especially in large systems with many
%interdependent components [10].''
%
% Referências do paper do Lanza para esse trecho:
%[8] M.-A. Storey, “Theories, methods and tools in program comprehension:
%Past, present and future,” in Proceedings of IWPC 2005. IEEE, 2005,
%pp. 181–191.
%[9] J. Sillito, G. C. Murphy, and K. De Volder, “Questions programmers
%ask during software evolution tasks,” in Proceedings FSE 2006. ACM,
%2006, pp. 23––34.
%[10] G. C. Murphy, D. Notkin, and K. Sullivan, “Software reflexion models:
%Bridging the gap between source and high-level models,” in Proceedings
%of FSE 1995. ACM, 1995, pp. 18–28.
\begin{highlight-box}{Takeaway T4 (Cognitive Load)}
  Duplications that made it possible to read the code without needing to jump
  to the function definition (to see its behavior) or the function call (to see
  its concrete parameters) were justifiable to reduce cognitive load when
  parsing through the code~\cite{skylines-cognitive-load}.
\end{highlight-box}

@@

\begin{highlight-box}{Takeaway T5 (Performance)}
  Given the low-level development context of Linux and the IIO subsystem, which
  is usually related to devices with limited resources, we detected hints that
  duplication may be forgivable if it improves performance.
\end{highlight-box}
